Bioverse: The Habitable Zone Inner Edge Discontinuity as an Imprint of Runaway Greenhouse Climates on Exoplanet Demographics

### 2023-07-07 src/tex/ms.tex
implement figure script for optimistic_RS_binned
implement figure script for M-R violins
add line numbers for PSJ submission
line numbers _and_ twocolumn for PSJ submission
avoid captures being cut
acknowledge referees

% Define document class
\documentclass[twocolumn,twocolappendix]{aastex631}
%\documentclass[modern]{aastex631}

@@

We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for sample sizes $\gtrsim$~100~planets if at least \SI{\sim 10}{\percent} of those orbiting closer than the habitable zone inner edge harbor runaway climates.
%We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for planet samples of size $\gtrsim$~100 if at least \SI{\sim 10}{\percent} of planets inside of the habitable zone inner edge harbor runaway climates.
Our survey simulations suggest that in the near future, ESA's \plato\ mission will be the most promising survey to probe the habitable zone inner edge discontinuity.
We determine survey strategies that maximize the diagnostic power of the obtained data and identify as key mission design drivers: 1. A follow-up campaign of planetary mass measurements and 2. The fraction of low-mass stars in the target sample.
Observational constraints on the runaway greenhouse transition will provide crucial insights into the distribution of atmospheric volatiles among rocky exoplanets, which may help to identify the nearest potentially habitable worlds.
\end{abstract}

@@

\subsection{Testability of the runaway greenhouse hypothesis}\label{sec:res_testability}
\begin{figure}[ht!]
    \script{optimistic_R-S.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S.pdf}
        \caption{
        Detection of the runaway greenhouse threshold.
        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
            The pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the habitable zone inner edge discontinuity.
Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is $f_\mathrm{rgh} = \frgh$, and we chose a water mass fraction of $x_{H_2O} = \var{wrr}$ for each planet.
In this case, the habitable zone inner edge discontinuity was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).

\begin{figure}[ht!]
    \script{cornerplot.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/corner.pdf}
        \caption{
            Retrieved posterior distribution of key parameters in the optimistic scenario. The density maps in each panel show relationships between and marginalized distributions of the threshold instellation $S_\mathrm{thresh}$, the water mass fraction $x_{H_2O}$, and the dilution factor $f_\mathrm{rgh}$ as they could be retrieved with a high-precision transit survey and a sample of \var{N_optimistic} planets. True values of the parameters for the injected effect are shown in orange. The threshold instellation can be reasonably constrained; the predominant water fraction and the fraction of planets with runaway greenhouse climates are degenerate.
        }
        \label{fig:cornerplot}
    \end{centering}
\end{figure}
With such a strong signal, we can attempt an inference of the parameters defining the injected effect.
Figure~\ref{fig:cornerplot} shows the posterior distributions of $S_\mathrm{thresh}$, $x_{H_2O}$, and $f_\mathrm{rgh}$ as determined by the nested sampler.
The threshold instellation can be accurately constrained with a median $\hat{S}_\mathrm{thresh} = \var{retrSthresh}$.
Both a higher water mass fraction and a higher dilution factor lead to larger average radii, thus these parameters are strongly correlated.

@@

\begin{acknowledgments}
The authors thank Gabriele Cugno, Yann Fabel, Brad Foley, Lisa Kaltenegger, Ravi Kopparapu, Eric Mamajek, Megan Mansfield, Paul Molliere, Gijs Mulders, Matthew Murphy, Sukrit Ranjan, and Terry-Ann Suer for insightful discussions.
We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
This material is based upon work supported by the National Aeronautics and Space Administration under Agreement No. 80NSSC21K0593 for the program ``Alien Earths''.
The results reported herein benefitted from collaborations and/or information exchange within NASA’s Nexus for Exoplanet System Science (NExSS) research coordination network sponsored by NASA’s Science Mission Directorate.
This work has made use of data from the European Space Agency (ESA) mission \gaia\ (\url{https://www.cosmos.esa.int/gaia}), processed by the \gaia\ Data Processing and Analysis Consortium (DPAC, \url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC has been provided by national institutions, in particular the institutions participating in the \gaia\ Multilateral Agreement.
T.L. was supported by a grant from the Branco Weiss Foundation.
G.B. acknowledges support from the NASA Astrophysics Data Analysis Program under Grant No.~80NSSC20K0446.
A.S. acknowledges support from NASA's Habitable Worlds Program (No.~80NSSC20K0226).
\end{acknowledgments}

@@

\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S_binned.pdf}
        \caption{
            Detection of the runaway greenhouse threshold with a binned statistic.
            Using simulated data from the optimistic survey case ($N = \var{N_optimistic}$, compare Figure~\ref{fig:optimistic_R-S}), we tested the runaway greenhouse hypothesis. Instead of using a running mean, we computed a binned statistic (arithmetic mean and standard deviation in blue) to assign planets the average radius or bulk density in their neighborhood in instellation space. Random draws from the posterior of the runaway greenhouse hypothesis (Eqn.~\ref{eq:rgh_hypo}) are shown in red.
            As in the nominal case, the pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S_binned}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S_binned} shows the simulated data together with binned, average planet radii and draws from the posterior of the hypothesis test.
A clear detection resulted, although with somewhat lower significance ($\Delta \ln \mathcal{Z} \approx 30$) compared to the nominal setup.
The accuracy of the recovered instellation threshold is comparable.
This test demonstrates that our results are not sensitive to the choice of statistical estimator to test the hypotheses against.

@@

\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/MR-violins.pdf}
        \caption{
        Posterior threshold instellations as a function of dilution factor for two alternative baseline mass-radius relations.
        For each grid step in $f_\mathrm{rgh}$, we show kernel density estimates of retrieved posteriors (averaged over 50 iterations) assuming an Earth-like (\SI{32.5}{\percent} Fe + \SI{67.5}{\percent} MgSiO$_3$) composition or the probabilistic relation from \citet{Wolfgang2016} and otherwise following the optimistic scenario in Sect.~\ref{sec:res_testability}.
        Lines within the violins show quartiles of the distributions, and the gray line indicates the injected threshold instellation of \var{injectedSthresh}.
        Higher log-evidence differences $\Delta \ln Z$ correspond to more significant rejections of the null hypothesis.
        In the regime of strong detections, both mass-radius relations lead to similar, accurate constraints on $S_\mathrm{thresh}$.
        Differences occur at low dilution factors, where the Earth-like relation leads to narrower estimates.
        }
        \label{fig:MR-violins}
    \end{centering}
\end{figure}
Figure~\ref{fig:MR-violins} shows how the two alternative baseline mass-radius relations influence the significance of a detection and the ability to constrain the threshold instellation.
The relation of \citet{Wolfgang2016} includes intrinsic scatter, which impedes a detection at low dilution factors.
In this regime, a layered, Earth-like composition leads to more significant detections and a narrower, although biased, constraint on $S_\mathrm{thresh}$.
Both mass-radius relations agree and recover the injected value where the null hypothesis can be rejected with high significance.
This is consistent with our nominal mass-radius relation (compare Sect.~\ref{sec:met-orbits_masses}).
We conclude that the underlying core and mantle composition of planets may affect the detectability of the transition if the fraction of planets with runaway greenhouse climates is low; however, the overall trends that our experiments revealed appear robust.

>>>>

% Define document class
%\documentclass[twocolumn,twocolappendix]{aastex631}
\documentclass[twocolumn,twocolappendix,linenumbers]{aastex631}
%\documentclass[modern]{aastex631}

@@

We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for sample sizes $\gtrsim$~100~planets if at least \SI{\sim 10}{\percent} of those orbiting closer than the habitable zone inner edge harbor runaway climates.
Our survey simulations suggest that in the near future, ESA's \plato\ mission will be the most promising survey to probe the habitable zone inner edge discontinuity.
We determine survey strategies that maximize the diagnostic power of the obtained data and identify as key mission design drivers: 1. A follow-up campaign of planetary mass measurements and 2. The fraction of low-mass stars in the target sample.
Observational constraints on the runaway greenhouse transition will provide crucial insights into the distribution of atmospheric volatiles among rocky exoplanets, which may help to identify the nearest potentially habitable worlds.
\end{abstract}

@@

\subsection{Testability of the runaway greenhouse hypothesis}\label{sec:res_testability}
\begin{figure}[ht!]
    \script{optimistic_RS.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S.pdf}
        \caption{
        Detection of the runaway greenhouse threshold.
        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
            The pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the habitable zone inner edge discontinuity.
Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is $f_\mathrm{rgh} = \frgh$, and we chose a water mass fraction of $x_{H_2O} = \var{wrr}$ for each planet.
In this case, the habitable zone inner edge discontinuity was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).

\begin{figure}[ht!]
    \script{cornerplot.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/corner.pdf}
        \caption{
            Retrieved posterior distribution of key parameters in the optimistic scenario. The density maps in each panel show relationships between and marginalized distributions of the threshold instellation $S_\mathrm{thresh}$, the water mass fraction $x_{H_2O}$, and the dilution factor $f_\mathrm{rgh}$ as they could be retrieved with a high-precision transit survey and a sample of \var{N_optimistic} planets. True values of the parameters for the injected effect are shown in orange. The threshold instellation can be reasonably constrained; the predominant water fraction and the fraction of planets with runaway greenhouse climates are degenerate.
        }
        \label{fig:cornerplot}
    \end{centering}
\end{figure}
With such a strong signal, we can attempt an inference of the parameters defining the injected effect.
Figure~\ref{fig:cornerplot} shows the posterior distributions of $S_\mathrm{thresh}$, $x_{H_2O}$, and $f_\mathrm{rgh}$ as determined by the nested sampler.
The threshold instellation can be accurately constrained. %with a median $\hat{S}_\mathrm{thresh} = \var{retrSthresh}$.
Both a higher water mass fraction and a higher dilution factor lead to larger average radii, thus these parameters are strongly correlated.

@@

\begin{acknowledgments}
We thank the anonymous referees for their constructive reports that helped to improve this manuscript.
The authors thank Gabriele Cugno, Yann Fabel, Brad Foley, Lisa Kaltenegger, Ravi Kopparapu, Eric Mamajek, Megan Mansfield, Paul Molliere, Gijs Mulders, Matthew Murphy, Sukrit Ranjan, and Terry-Ann Suer for insightful discussions.
We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
This material is based upon work supported by the National Aeronautics and Space Administration under Agreement No. 80NSSC21K0593 for the program ``Alien Earths''.
The results reported herein benefitted from collaborations and/or information exchange within NASA’s Nexus for Exoplanet System Science (NExSS) research coordination network sponsored by NASA’s Science Mission Directorate.
This work has made use of data from the European Space Agency (ESA) mission \gaia\ (\url{https://www.cosmos.esa.int/gaia}), processed by the \gaia\ Data Processing and Analysis Consortium (DPAC, \url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC has been provided by national institutions, in particular the institutions participating in the \gaia\ Multilateral Agreement.
T.L. was supported by a grant from the Branco Weiss Foundation.
G.B. acknowledges support from the NASA Astrophysics Data Analysis Program under Grant No.~80NSSC20K0446.
A.S. acknowledges support from NASA's Habitable Worlds Program (No.~80NSSC20K0226).
\end{acknowledgments}

@@

\begin{figure}[ht!]
    \script{optimistic_RS_binned.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S_binned.pdf}
        \caption{
            Detection of the runaway greenhouse threshold with a binned statistic.
            Using simulated data from the optimistic survey case ($N = \var{N_optimistic}$, compare Figure~\ref{fig:optimistic_R-S}), we tested the runaway greenhouse hypothesis. Instead of using a running mean, we computed a binned statistic (arithmetic mean and standard deviation in blue) to assign planets the average radius or bulk density in their neighborhood in instellation space. Random draws from the posterior of the runaway greenhouse hypothesis (Eqn.~\ref{eq:rgh_hypo}) are shown in red.
            As in the nominal case, the pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S_binned}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S_binned} shows the simulated data together with binned, average planet radii and draws from the posterior of the hypothesis test.
A clear detection resulted, although with somewhat lower significance ($\Delta \ln \mathcal{Z} \approx 30$) compared to the nominal setup.
The accuracy of the recovered instellation threshold is comparable.
This test demonstrates that our results are not sensitive to the choice of statistical estimator to test the hypotheses against.

@@

\begin{figure}[ht!]
    \script{MR_violins.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/MR-violins.pdf}
        \caption{
        Posterior threshold instellations as a function of dilution factor for two alternative baseline mass-radius relations.
        For each grid step in $f_\mathrm{rgh}$, we show kernel density estimates of retrieved posteriors (averaged over 50 iterations) assuming an Earth-like (\SI{32.5}{\percent} Fe + \SI{67.5}{\percent} MgSiO$_3$) composition or the probabilistic relation from \citet{Wolfgang2016} and otherwise following the optimistic scenario in Sect.~\ref{sec:res_testability}.
        Lines within the violins show quartiles of the distributions, and the gray line indicates the injected threshold instellation of \var{injectedSthresh}.
        Higher log-evidence differences $\Delta \ln Z$ correspond to more significant rejections of the null hypothesis.
        In the regime of strong detections, both mass-radius relations lead to similar, accurate constraints on $S_\mathrm{thresh}$.
        Differences occur at low dilution factors, where the Earth-like relation leads to narrower estimates.
        }
        \label{fig:MR-violins}
    \end{centering}
\end{figure}
Figure~\ref{fig:MR-violins} shows how the two alternative baseline mass-radius relations influence the significance of a detection and the ability to constrain the threshold instellation.
The relation of \citet{Wolfgang2016} includes intrinsic scatter, which impedes a detection at low dilution factors.
In this regime, a layered, Earth-like composition leads to more significant detections and a narrower, although biased, constraint on $S_\mathrm{thresh}$.
Both mass-radius relations agree and recover the injected value where the null hypothesis can be rejected with high significance.
This is consistent with our nominal mass-radius relation (compare Sect.~\ref{sec:met-orbits_masses}).
We conclude that the underlying core and mantle composition of planets may affect the detectability of the transition if the fraction of planets with runaway greenhouse climates is low; however, the overall trends that our experiments revealed appear robust.
